---
title: "Final Report"
author: "Liz Kim"
date: "1/21/2025"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(rvest)
library(tidyverse)
library(dplyr)
library(tidyr)
library(readr)
library(knitr) 
```

## Table of Contents

-   [1. Intro](#intro)
-   [2. Research Question](#research-question)
-   [3. Data Scraping & Tidying the Data](#Data-Scaping-and-Tidying-the-Data)
-   [4. Descriptive Statistics and Plots](#descriptive-statistics-and-plots)
-   [5. Data Interpretation](#data-interpretation)

# 1. Introduction {#introduction}

This project investigates the factors associated with national happiness, utilizing happiness scores from the World Population Review and key factors from Wikipedia tables. Key variables include GDP per capita, environmental quality, abd literacy rates offering insights into the economic, environmental, and educational factors that influence happiness levels across countries.

The reason why I'm working on this project is because it is important to understand how these factors contribute to national happiness. Happiness is a critical measure of well-being, and identifying its determinants is vital. abcdedfdfadfadfd

# 2. Research Question {#research-question}

Main Question: How do economic,education, and environmental variables influence national happiness levels as reported by the World Population Review?

RQ1: How does GDP per capita associate with a country's happiness score?
RQ2: How does environmental quality associate with a country's happiness score?
RQ3: How does a country's literacy rate associate with its happiness score?

# 3. Data Scraping & Tidying the Data

Below is the code I used to scrap the data. 

**Happiness Score**

```{r}
# Read the CSV file
happiest_countries <- read_csv("happiest-countries-in-the-world-2024.csv")

# Preview the data
head(happiest_countries)
```

I requested this data table from WorldPopulationReview.com. I received an excel file through my email.


**GDP per Capita**

```{r}
# Save the URL and scrape the webpage
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)_per_capita"
gdp_page <- read_html(x = url)

# Extract all tables with the class "wikitable"
tables <- html_elements(gdp_page, css = "table.wikitable")

# Convert the first table to a data frame
gdp_per_capita_table <- html_table(tables[[1]], fill = TRUE)

# Save the table as a CSV file
write_csv(gdp_per_capita_table, "gdp_per_capita.csv")

# Preview the table
head(gdp_per_capita_table)
```

**Education (Literacy Rate)**

```{r}
# Save the URL and scrape the webpage
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_literacy_rate"
literacy_page <- read_html(x = url)

# Extract all tables with the class "wikitable"
tables <- html_elements(literacy_page, css = "table.wikitable")

# Convert all table nodes into a list of data frames
all_tables <- html_table(tables, fill = TRUE)

# Inspect tables to find the one I want to extract
for (i in seq_along(all_tables)) {
  print(paste("Table", i))
  print(head(all_tables[[i]]))
}

# Extract the second table 
literacy_rate_table <- all_tables[[2]]
write_csv(literacy_rate_table, "literacy_rate.csv")

# Preview the table
head(literacy_rate_table)
```

**Environmental Quality**

```{r}
# Save the URL and scrape the webpage
url <- "https://en.wikipedia.org/wiki/Environmental_Performance_Index"
environment_page <- read_html(x = url)

# Extract all tables with the class "wikitable"
tables <- html_elements(environment_page, css = "table.wikitable")

# Convert all table nodes into a list of data frames
all_tables <- html_table(tables, fill = TRUE)

# Inspect tables to find the one I want to extract
for (i in seq_along(all_tables)) {
  print(paste("Table", i))
  print(head(all_tables[[i]]))
}

# Extract the second table (Environmental Performance Index)
environmental_quality_table <- all_tables[[2]]
write_csv(environmental_quality_table, "environmental_quality.csv")

# Preview the table
head(environmental_quality_table)
```

Here, I included the code to tidy the tables. The datasets comprises four tables, each offering insights into countries' happiness scores, GDP, literacy rates, and environmental performance. To briefly define “tidy” data and explain its advantages and disadvantages, 

## Happiness Score Table

```{r}
# Read the previously saved CSV file
happiest_countries <- read_csv("happiest-countries-in-the-world-2024.csv")

# Clean and rename the happiest_countries dataset
happiest_countries_cleaned <- happiest_countries %>%
  select(country, HappiestCountriesWorldHappinessReportScore2024) %>% 
  filter(!is.na(HappiestCountriesWorldHappinessReportScore2024)) %>% 
  rename(HappinessScore = HappiestCountriesWorldHappinessReportScore2024) 

# Preview and save the cleaned table
head(happiest_countries_cleaned)
write_csv(happiest_countries_cleaned, "happiest_countries_cleaned.csv")
```


## GDP Table

```{r}
# Read the previously saved CSV file
gdp_per_capita_table <- read_csv("gdp_per_capita.csv")

# Rename columns
colnames(gdp_per_capita_table) <- c("country", "IMF_estimate", "IMF_year", 
                                    "World_Bank_estimate", "World_Bank_year", 
                                    "UN_estimate", "UN_year")

# Tidy the table
fill_missing_values <- function(data) {
  data %>%
    select(country, IMF_estimate) %>%              
    mutate(
      IMF_estimate = ifelse(IMF_estimate == "-", NA, IMF_estimate), 
      IMF_estimate = gsub(",", "", IMF_estimate),      
      IMF_estimate = as.numeric(IMF_estimate)         
    ) %>%
    fill(IMF_estimate, .direction = "down") %>%       
    filter(!is.na(IMF_estimate)) %>%                  
    rename(estimate = IMF_estimate)                  
}

# Apply the cleaning function
gdp_per_capita_table_cleaned <- fill_missing_values(gdp_per_capita_table)

# Preview and save the cleaned table
head(gdp_per_capita_table_cleaned) 
write_csv(gdp_per_capita_table_cleaned, "gdp_per_capita_table_cleaned.csv")
```


## Education (Literacy Rate Table)

```{r}
# Read the previously saved CSV file
literacy_rate_table <- read_csv("literacy_rate.csv")
head(literacy_rate_table)

# Rename columns
colnames(literacy_rate_table) <- c("country", "total", "male", "female", "gap", "year")

# Tidy the data
literacy_rate_table_cleaned <- literacy_rate_table %>%
  select(country, total, year) %>%
  mutate(
    country = gsub("\\*", "", country),  
    total = as.numeric(gsub("%", "", total)), 
    year = as.numeric(year)                 
  )

# Filter out rows where country == "Country" or "World"
literacy_rate_table_cleaned <- literacy_rate_table_cleaned %>%
  filter(country != "Country",              
         country != "World")    

# Preview and save the cleaned table
head(literacy_rate_table_cleaned)
write_csv(literacy_rate_table_cleaned, "literacy_rate_table_cleaned.csv")
```


## Environmental Quality Table

```{r}
# Read the previously saved CSV file
environmental_quality_table <- read_csv("environmental_quality.csv")

# Tidy the data
environmental_quality_table_cleaned <- environmental_quality_table %>%
  select(country = `Country`, value = `Value`)

# Preview and save the cleaned table
head(environmental_quality_table_cleaned)
write_csv(environmental_quality_table_cleaned, "environmental_quality_table_cleaned.csv")
```

# 4. Descriptive Statistics and Plots {#descriptive-statistics-and-plots}

```{r}
# Clean country names function
clean_country_names <- function(data, country_column = "country") {
  data %>%
    mutate(!!sym(country_column) := gsub("[^[:alnum:] ]", "", !!sym(country_column))) %>%
    mutate(!!sym(country_column) := trimws(!!sym(country_column))) %>%
    mutate(!!sym(country_column) := tolower(!!sym(country_column)))
}

# Clean the country names in all datasets
happiest_countries_cleaned <- clean_country_names(happiest_countries_cleaned)
gdp_per_capita_table_cleaned <- clean_country_names(gdp_per_capita_table_cleaned)
literacy_rate_table_cleaned <- clean_country_names(literacy_rate_table_cleaned)
environmental_quality_table_cleaned <- clean_country_names(environmental_quality_table_cleaned)

# Rename columns to prepare for merging
gdp_data <- gdp_per_capita_table_cleaned %>%
  rename(Value = estimate) %>%
  mutate(Variable = "GDP")

education_data <- literacy_rate_table_cleaned %>%
  rename(Value = total) %>%
  mutate(Variable = "Education")

env_quality_data <- environmental_quality_table_cleaned %>%
  rename(Value = value) %>%
  mutate(Variable = "EnvironmentalQuality")

happiness_data <- happiest_countries_cleaned %>%
  rename(Value = HappinessScore) %>%
  mutate(Variable = "HappinessScore")

# Combine all datasets into a single long format
long_data <- bind_rows(
  gdp_data,
  education_data,
  env_quality_data,
  happiness_data
)

# Pivot to wide format
wide_data <- long_data %>%
  pivot_wider(
    names_from = Variable,
    values_from = Value
  )

# Remove year and sexgap columns
wide_data_cleaned <- wide_data %>%
  select(-year) 

# Display the Merged Data Overview
wide_data_cleaned %>%
  head(10) %>% 
  kable(
    col.names = c("Country", "GDP", "Education", "Environmental Quality", "Happiness Score"),
    caption = "Merged Data Overview",
    format = "markdown"
  )
```

I created the Merged Data Overview table to provide a comprehensive view of all the important variables in a single table. However, due to differences in country names and variations in the countries included in each dataset, many NA values are present. Therefore, for the visualizations, I will analyze two variables at a time separately. Despite the NA values, this table is still useful for quickly skimming through the data.

## RQ1: How does GDP per capita associate with a country's happiness score?

```{r}
# Clean the country names in gdp_per_capita_table_cleaned
gdp_per_capita_table_cleaned <- gdp_per_capita_table_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                   
  mutate(country = tolower(country))                      

# Clean the country names in happiest_countries_cleaned
happiest_countries_cleaned <- happiest_countries_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                    
  mutate(country = tolower(country))                     

# Merge GDP and HappinessScore data
merged_data <- gdp_per_capita_table_cleaned %>%
  left_join(happiest_countries_cleaned, by = "country")
print(merged_data)

# Remove missing values
cleaned_merged_data <- merged_data %>%
  filter(!is.na(estimate), !is.na(HappinessScore))
print(cleaned_merged_data)

# View summary statistics 
summary(cleaned_merged_data[, c("estimate", "HappinessScore")])

# Create the scatter plot
ggplot(data = cleaned_merged_data, aes(x = estimate, y = HappinessScore)) +
    geom_point(color = "blue", size = 2) +                
    geom_smooth(method = "lm", color = "red", se = TRUE) + 
    scale_x_continuous(labels = scales::comma) +          
    labs(
        x = "GDP Per Capita (USD)",
        y = "Happiness Score",
        title = "GDP Per Capita and Happiness Score"
    ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )
```

## RQ2: How does environmental quality associate with a country's happiness score?

```{r}
# Clean the country names in environmental_quality_table_cleaned
environmental_quality_table_cleaned <- environmental_quality_table_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                  
  mutate(country = tolower(country))                     

# Clean the country names in happiest_countries_cleaned
happiest_countries_cleaned <- happiest_countries_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                    
  mutate(country = tolower(country))                     

# Merge environmental quality data with happiness data
merged_data_env <- environmental_quality_table_cleaned %>%
  left_join(happiest_countries_cleaned, by = "country")
print(merged_data_env)

# Remove rows with missing values in value or HappinessScore
cleaned_merged_data_env <- merged_data_env %>%
  filter(!is.na(value), !is.na(HappinessScore))
print(cleaned_merged_data_env)

# View summary statistics 
summary(cleaned_merged_data_env[, c("value")])

# Create the line graph
ggplot(cleaned_merged_data_env, aes(x = value, y = HappinessScore)) +
  geom_line(color = "blue", size = 1) +  
  geom_point(color = "red", size = 2) + 
  labs(
    title = "Environmental Index and Happiness Score",
    x = "Environmental Index",
    y = "Happiness Score"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )

# Create the scatter plot
ggplot(cleaned_merged_data_env, aes(x = value, y = HappinessScore)) +
  geom_point(color = "blue", size = 2) +               
  geom_smooth(method = "lm", color = "red", se = TRUE) + 
  labs(
    title = "Environmental Index and Happiness Score",
    x = "Environmental Index",
    y = "Happiness Score"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )
```

I created two graphs to explore and to find out the best graph. The line graph still helps demonstrate a general positive trend between environmental quality and happiness scores; however, I think the scatter plot graph is better.

## RQ3: How does a country's literacy rate associate with its happiness score?

```{r}
# Clean the country names in literacy_rate_table_cleaned
literacy_rate_table_cleaned <- literacy_rate_table_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>% 
  mutate(country = trimws(country)) %>%                    
  mutate(country = tolower(country))                    

# Clean the country names in happiest_countries_cleaned
happiest_countries_cleaned <- happiest_countries_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>% 
  mutate(country = trimws(country)) %>%                   
  mutate(country = tolower(country))                    

# Merge Literacy Rate and HappinessScore data
merged_data_lit <- literacy_rate_table_cleaned %>%
  left_join(happiest_countries_cleaned, by = "country")
print(merged_data_lit)

# Remove missing values
cleaned_merged_data_lit <- merged_data_lit %>%
  filter(!is.na(total), !is.na(year), !is.na(HappinessScore))
print(cleaned_merged_data_lit)

# View summary statistics 
summary(cleaned_merged_data_lit[, c("total")])

# Calculate percentiles for the "total" column in the literacy rate table
percentiles <- quantile(literacy_rate_table_cleaned$total, probs = c(0.25, 0.5, 0.75, 1), na.rm = TRUE)
print(percentiles)

# Add bins to the data using the specified quartiles
cleaned_merged_data_lit <- cleaned_merged_data_lit %>%
  mutate(
    literacy_bin = cut(
      total,
      breaks = c(0, 80.725, 95.650, 99.000, 100), 
      labels = c("Q1: <80.725", 
                 "Q2: 80.725-95.650", 
                 "Q3: 95.650-99.000", 
                 "Q4: >99.000"),
      include.lowest = TRUE
    )
  )
# Create the box plot
ggplot(cleaned_merged_data_lit, aes(x = literacy_bin, y = HappinessScore)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.color = "red", outlier.size = 2) +
  labs(
    title = "Literacy Rate and Happiness Score",
    x = "Literacy Rate (Quartiles)",
    y = "Happiness Score"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )
```
# 5. Data Interpretation {#data-interpretation}

## Data Analysis

In addressing the main question, "How do economic, environmental, and educational variables influence national happiness levels as reported by the World Population Review?", this analysis provides evidence of significant associations between these factors and happiness scores.

*H1: I hypothesize that higher GDP per capita, measured in USD based on IMF estimates, is associated with higher national happiness levels.*

The scatter plot examining GDP per capita and happiness scores reveals a clear positive trend. Countries with higher GDP per capita tend to report higher happiness scores. This supports the hypothesis that wealthier countries have more resources to invest in public services such as healthcare, education, and infrastructure, which improve the overall quality of life.

*H2: I hypothesize that better environmental quality, as measured by the Environmental Performance Index (EPI), is associated with higher happiness.*

The line graph analyzing the association between environmental quality and happiness scores suggests a positive but variable association. Countries with higher environmental index scores generally report higher happiness levels, supporting the hypothesis that a clean and sustainable environment promotes well-being. However, the noisy pattern in the graph indicates that some countries with similar environmental scores differ significantly in happiness. The scatter plot graph shows a positive trend as well.

*H3: I hypothesize that higher literacy rates, as a measure of education, are associated with higher happiness scores.*

The box plot demonstrates a generally positive association between literacy rates and happiness scores across the first three quartiles (Q1 to Q3). However, Q3 (95.650--99.000) appears to have a slightly higher happiness scores compared to Q4 (\> 99.000), despite Q4 representing the countries with the highest literacy rates. This pattern could reflect diminishing returns of literacy on happiness or challenges unique to highly literate societies, such as increased societal pressures, disparities in resource allocation, or inequality.

## Limitations and Future Directions

One limitation of this study lies in the quality and consistency of the data used. Data sources may vary in their methods of collection, measurements, and reporting standards across countries. Future research should prioritize using standardized global datasets to minimize inconsistencies and align data collection periods across variables to ensure accuracy. Conducting longitudinal studies could further enhance the understanding of how changes in variables influence happiness over time, providing a more comprehensive and precise perspective on these relationships.

As an extension of this study, we could perform statistical correlation tests, such as Pearson or Spearman correlation, to quantitatively assess the strength and direction of relationships between each variable and national happiness scores. This approach would allow us to identify which factor is most strongly correlated with happiness and could potentially be a primary determinant. Additionally, conducting regression analysis could help control for confounding variables and provide deeper insights into the relative importance of each factor in explaining variations in happiness.

## Implications

The positive link between GDP per capita and happiness emphasizes the importance of investments in public services like healthcare, education, and infrastructure, with a focus on reducing income inequality to maximize benefits. Similarly, the association between environmental quality and happiness underscores the need for sustainable policies that prioritize pollution reduction and green initiatives. The relationship between literacy rates and happiness suggests that education policies should focus not only on improving literacy but also on ensuring equitable access and opportunities, as the trend indicates that higher literacy rates do not always directly translate to higher happiness scores. Overall, these findings demonstrate the need for balanced strategies addressing economic, educational, and environmental factors to achieve sustainable well-being globally.
