---
title: "Final Report"
author: "Liz Kim"
date: "1/21/2025"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(rvest)
library(tidyverse)
library(dplyr)
library(tidyr)
library(readr)
library(knitr) 
```

## Table of Contents

-   [1. Introduction](#introduction)
-   [2. Research Question](#research-question)
-   [3. Data Scraping & Tidying the Data](#Data-Scaping-and-Tidying-the-Data)
-   [4. Descriptive Statistics and Plots](#descriptive-statistics-and-plots)
-   [5. Data Interpretation](#data-interpretation)

# 1. Introduction {#introduction}

This project investigates the factors associated with national happiness, utilizing happiness scores from the World Population Review and key factors from Wikipedia tables. Key variables include GDP per capita, environmental quality, abd literacy rates offering insights into the economic, environmental, and educational factors that influence happiness levels across countries.

The reason why I'm working on this project is because it is important to understand how these factors contribute to national happiness. Happiness is a critical measure of well-being, and identifying its determinants is vital.

# 2. Research Question {#research-question}

How do economic,education, and environmental variables influence national happiness levels as reported by the World Population Review?

# 3. Data Scraping & Tidying the Data

The codes for data scraping are provided in a separate file: DataScraping.rmd. In this section, I included the code to tidy the tables and provided brief data descriptions. The datasets comprises five tables, each offering insights into countries' happiness scores, GDP, literacy rates, environmental performance, and life expectancy.

**Happiness Score**

```{r}
# Read the CSV file
happiest_countries <- read_csv("happiest-countries-in-the-world-2024.csv")

# Preview the data
head(happiest_countries)
```

I requested this data table from WorldPopulationReview.com. I received an excel file through my email.


**GDP per Capita**

```{r}
# Save the URL and scrape the webpage
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(nominal)_per_capita"
gdp_page <- read_html(x = url)

# Extract all tables with the class "wikitable"
tables <- html_elements(gdp_page, css = "table.wikitable")

# Convert the first table to a data frame
gdp_per_capita_table <- html_table(tables[[1]], fill = TRUE)

# Save the table as a CSV file
write_csv(gdp_per_capita_table, "gdp_per_capita.csv")

# Preview the table
head(gdp_per_capita_table)
```

**Education (Literacy Rate)**

```{r}
# Save the URL and scrape the webpage
url <- "https://en.wikipedia.org/wiki/List_of_countries_by_literacy_rate"
literacy_page <- read_html(x = url)

# Extract all tables with the class "wikitable"
tables <- html_elements(literacy_page, css = "table.wikitable")

# Convert all table nodes into a list of data frames
all_tables <- html_table(tables, fill = TRUE)

# Inspect tables to find the one I want to extract
for (i in seq_along(all_tables)) {
  print(paste("Table", i))
  print(head(all_tables[[i]]))
}

# Extract the second table 
literacy_rate_table <- all_tables[[2]]
write_csv(literacy_rate_table, "literacy_rate.csv")

# Preview the table
head(literacy_rate_table)
```

**Environmental Quality**

```{r}
# Save the URL and scrape the webpage
url <- "https://en.wikipedia.org/wiki/Environmental_Performance_Index"
environment_page <- read_html(x = url)

# Extract all tables with the class "wikitable"
tables <- html_elements(environment_page, css = "table.wikitable")

# Convert all table nodes into a list of data frames
all_tables <- html_table(tables, fill = TRUE)

# Inspect tables to find the one I want to extract
for (i in seq_along(all_tables)) {
  print(paste("Table", i))
  print(head(all_tables[[i]]))
}

# Extract the second table (Environmental Performance Index)
environmental_quality_table <- all_tables[[2]]
write_csv(environmental_quality_table, "environmental_quality.csv")

# Preview the table
head(environmental_quality_table)
```

Here, I included the code to tidy the tables and provided brief data descriptions. The datasets comprises four tables, each offering insights into countries' happiness scores, GDP, literacy rates, and environmental performance.

## Happiness Score Table

```{r}
# Read the previously saved CSV file
happiest_countries <- read_csv("happiest-countries-in-the-world-2024.csv")

# Clean and rename the happiest_countries dataset
happiest_countries_cleaned <- happiest_countries %>%
  select(country, HappiestCountriesWorldHappinessReportScore2024) %>% 
  filter(!is.na(HappiestCountriesWorldHappinessReportScore2024)) %>% 
  rename(HappinessScore = HappiestCountriesWorldHappinessReportScore2024) 

# Preview and save the cleaned table
head(happiest_countries_cleaned)
write_csv(happiest_countries_cleaned, "happiest_countries_cleaned.csv")
```


## GDP Table

```{r}
# Read the previously saved CSV file
gdp_per_capita_table <- read_csv("gdp_per_capita.csv")

# Rename columns
colnames(gdp_per_capita_table) <- c("country", "IMF_estimate", "IMF_year", 
                                    "World_Bank_estimate", "World_Bank_year", 
                                    "UN_estimate", "UN_year")

# Tidy the table
fill_missing_values <- function(data) {
  data %>%
    select(country, IMF_estimate) %>%              
    mutate(
      IMF_estimate = ifelse(IMF_estimate == "-", NA, IMF_estimate), 
      IMF_estimate = gsub(",", "", IMF_estimate),      
      IMF_estimate = as.numeric(IMF_estimate)         
    ) %>%
    fill(IMF_estimate, .direction = "down") %>%       
    filter(!is.na(IMF_estimate)) %>%                  
    rename(estimate = IMF_estimate)                  
}

# Apply the cleaning function
gdp_per_capita_table_cleaned <- fill_missing_values(gdp_per_capita_table)

# Preview and save the cleaned table
head(gdp_per_capita_table_cleaned) 
write_csv(gdp_per_capita_table_cleaned, "gdp_per_capita_table_cleaned.csv")
```


## Education (Literacy Rate Table)

```{r}
# Read the previously saved CSV file
literacy_rate_table <- read_csv("literacy_rate.csv")
head(literacy_rate_table)

# Rename columns
colnames(literacy_rate_table) <- c("country", "total", "male", "female", "gap", "year")

# Tidy the data
literacy_rate_table_cleaned <- literacy_rate_table %>%
  select(country, total, year) %>%
  mutate(
    country = gsub("\\*", "", country),  
    total = as.numeric(gsub("%", "", total)), 
    year = as.numeric(year)                 
  )

# Filter out rows where country == "Country" or "World"
literacy_rate_table_cleaned <- literacy_rate_table_cleaned %>%
  filter(country != "Country",              
         country != "World")    

# Preview and save the cleaned table
head(literacy_rate_table_cleaned)
write_csv(literacy_rate_table_cleaned, "literacy_rate_table_cleaned.csv")
```


## Environmental Quality Table

```{r}
# Read the previously saved CSV file
environmental_quality_table <- read_csv("environmental_quality.csv")

# Tidy the data
environmental_quality_table_cleaned <- environmental_quality_table %>%
  select(country = `Country`, value = `Value`)

# Preview and save the cleaned table
head(environmental_quality_table_cleaned)
write_csv(environmental_quality_table_cleaned, "environmental_quality_table_cleaned.csv")
```

# 4. Descriptive Statistics and Plots {#descriptive-statistics-and-plots}

```{r}
# Clean country names function
clean_country_names <- function(data, country_column = "country") {
  data %>%
    mutate(!!sym(country_column) := gsub("[^[:alnum:] ]", "", !!sym(country_column))) %>%
    mutate(!!sym(country_column) := trimws(!!sym(country_column))) %>%
    mutate(!!sym(country_column) := tolower(!!sym(country_column)))
}

# Clean the country names in all datasets
happiest_countries_cleaned <- clean_country_names(happiest_countries_cleaned)
gdp_per_capita_table_cleaned <- clean_country_names(gdp_per_capita_table_cleaned)
literacy_rate_table_cleaned <- clean_country_names(literacy_rate_table_cleaned)
environmental_quality_table_cleaned <- clean_country_names(environmental_quality_table_cleaned)

# Rename columns to prepare for merging
gdp_data <- gdp_per_capita_table_cleaned %>%
  rename(Value = estimate) %>%
  mutate(Variable = "GDP")

education_data <- literacy_rate_table_cleaned %>%
  rename(Value = total) %>%
  mutate(Variable = "Education")

env_quality_data <- environmental_quality_table_cleaned %>%
  rename(Value = value) %>%
  mutate(Variable = "EnvironmentalQuality")

happiness_data <- happiest_countries_cleaned %>%
  rename(Value = HappinessScore) %>%
  mutate(Variable = "HappinessScore")

# Combine all datasets into a single long format
long_data <- bind_rows(
  gdp_data,
  education_data,
  env_quality_data,
  happiness_data
)

# Pivot to wide format
wide_data <- long_data %>%
  pivot_wider(
    names_from = Variable,
    values_from = Value
  )

# Remove year and sexgap columns
wide_data_cleaned <- wide_data %>%
  select(-year) 

# Display the Merged Data Overview
wide_data_cleaned %>%
  head(10) %>% 
  kable(
    col.names = c("Country", "GDP", "Education", "Environmental Quality", "Happiness Score"),
    caption = "Merged Data Overview",
    format = "markdown"
  )
```

I created the Merged Data Overview table to provide a comprehensive view of all the important variables in a single table. However, due to differences in country names and variations in the countries included in each dataset, many NA values are present. Therefore, for the visualizations, I will analyze two variables at a time separately. Despite the NA values, this table is still useful for quickly skimming through the data.

## RQ1: How does GDP per capita associate with a country's happiness score?

```{r}
# Clean the country names in gdp_per_capita_table_cleaned
gdp_per_capita_table_cleaned <- gdp_per_capita_table_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                   
  mutate(country = tolower(country))                      

# Clean the country names in happiest_countries_cleaned
happiest_countries_cleaned <- happiest_countries_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                    
  mutate(country = tolower(country))                     

# Merge GDP and HappinessScore data
merged_data <- gdp_per_capita_table_cleaned %>%
  left_join(happiest_countries_cleaned, by = "country")
print(merged_data)

# Remove missing values
cleaned_merged_data <- merged_data %>%
  filter(!is.na(estimate), !is.na(HappinessScore))
print(cleaned_merged_data)

# View summary statistics 
summary(cleaned_merged_data[, c("estimate", "HappinessScore")])

# Create the scatter plot
ggplot(data = cleaned_merged_data, aes(x = estimate, y = HappinessScore)) +
    geom_point(color = "blue", size = 2) +                
    geom_smooth(method = "lm", color = "red", se = TRUE) + 
    scale_x_continuous(labels = scales::comma) +          
    labs(
        x = "GDP Per Capita (USD)",
        y = "Happiness Score",
        title = "GDP Per Capita and Happiness Score"
    ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )
```

## RQ2: How does environmental quality associate with a country's happiness score?

```{r}
# Clean the country names in environmental_quality_table_cleaned
environmental_quality_table_cleaned <- environmental_quality_table_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                  
  mutate(country = tolower(country))                     

# Clean the country names in happiest_countries_cleaned
happiest_countries_cleaned <- happiest_countries_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>%  
  mutate(country = trimws(country)) %>%                    
  mutate(country = tolower(country))                     

# Merge environmental quality data with happiness data
merged_data_env <- environmental_quality_table_cleaned %>%
  left_join(happiest_countries_cleaned, by = "country")
print(merged_data_env)

# Remove rows with missing values in value or HappinessScore
cleaned_merged_data_env <- merged_data_env %>%
  filter(!is.na(value), !is.na(HappinessScore))
print(cleaned_merged_data_env)

# View summary statistics 
summary(cleaned_merged_data_env[, c("value")])

# Create the line graph
ggplot(cleaned_merged_data_env, aes(x = value, y = HappinessScore)) +
  geom_line(color = "blue", size = 1) +  
  geom_point(color = "red", size = 2) + 
  labs(
    title = "Environmental Index and Happiness Score",
    x = "Environmental Index",
    y = "Happiness Score"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )

# Create the scatter plot
ggplot(cleaned_merged_data_env, aes(x = value, y = HappinessScore)) +
  geom_point(color = "blue", size = 2) +               
  geom_smooth(method = "lm", color = "red", se = TRUE) + 
  labs(
    title = "Environmental Index and Happiness Score",
    x = "Environmental Index",
    y = "Happiness Score"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )
```

I created two graphs to explore and to find out the best graph. The line graph still helps demonstrate a general positive trend between environmental quality and happiness scores; however, I think the scatter plot graph is better.

## RQ3: How does a country's literacy rate associate with its happiness score?

```{r}
# Clean the country names in literacy_rate_table_cleaned
literacy_rate_table_cleaned <- literacy_rate_table_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>% 
  mutate(country = trimws(country)) %>%                    
  mutate(country = tolower(country))                    

# Clean the country names in happiest_countries_cleaned
happiest_countries_cleaned <- happiest_countries_cleaned %>%
  mutate(country = gsub("[^[:alnum:] ]", "", country)) %>% 
  mutate(country = trimws(country)) %>%                   
  mutate(country = tolower(country))                    

# Merge Literacy Rate and HappinessScore data
merged_data_lit <- literacy_rate_table_cleaned %>%
  left_join(happiest_countries_cleaned, by = "country")
print(merged_data_lit)

# Remove missing values
cleaned_merged_data_lit <- merged_data_lit %>%
  filter(!is.na(total), !is.na(year), !is.na(HappinessScore))
print(cleaned_merged_data_lit)

# View summary statistics 
summary(cleaned_merged_data_lit[, c("total")])

# Calculate percentiles for the "total" column in the literacy rate table
percentiles <- quantile(literacy_rate_table_cleaned$total, probs = c(0.25, 0.5, 0.75, 1), na.rm = TRUE)
print(percentiles)

# Add bins to the data using the specified quartiles
cleaned_merged_data_lit <- cleaned_merged_data_lit %>%
  mutate(
    literacy_bin = cut(
      total,
      breaks = c(0, 80.725, 95.650, 99.000, 100), 
      labels = c("Q1: <80.725", 
                 "Q2: 80.725-95.650", 
                 "Q3: 95.650-99.000", 
                 "Q4: >99.000"),
      include.lowest = TRUE
    )
  )
# Create the box plot
ggplot(cleaned_merged_data_lit, aes(x = literacy_bin, y = HappinessScore)) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.color = "red", outlier.size = 2) +
  labs(
    title = "Literacy Rate and Happiness Score",
    x = "Literacy Rate (Quartiles)",
    y = "Happiness Score"
  ) +
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 12) 
  )
```
